Human Factors Influencing Visual Statistical Inference

### 2014-10-26 sociology_paper.tex
edits factors

In the initial Turk experiments (1,2,3) conducted with simulated data to compare the results from visual inference with those obtained by classical inference, the questions posed to the observers were more targeted. This was important \hh{to allow a direct comparison of the results to results from conventional testing.}%in order to make a comparison of the results. 
For latter studies (4, 7, 10), questions posed to the observer were more general. Table \ref{tbl:visual_stat} lists the questions used for each experiment.

@@

%\subsection{Learning Trend} 
\item{\bf Learning Trend:} One might expect that as an observer evaluates more lineups they may become more skillful in their evaluations. Each subject in the experiments 5, 6 and 7 evaluated a block of of ten lineups of the same type of data plot. The ten lineups were randomly chosen from the lineups  produced for each study. Before evaluating the first lineup, the observer needs to read instructions and become accustomed to the type of plot used in the lineup. In subsequent lineups being evaluated the type of plot is the same, level of difficulty will be different. It is possible that the observer becomes more skillful at recognizing the most different plot, either by more often detecting the actual data plot or more quickly reporting their choice. These two ways of measuring learning trend are evaluated on data collected from experiments 5, 6 and 7.

@@

%\subsection{Location of Actual Data Plot in the Lineup}
\item{\bf Location of Actual Data Plot in the Lineup} For all of the lineups used in the experiments a $5\times 4$ grid of 20 points is used. A random number generator is used to determine the position where the actual data plot is placed in each lineup. With the help of eye-tracking equipments \citet{zhao:2012} tracked the observers' eyes to see how they searched the plots in a lineup. The results suggest that many people read lineups from left to right direction, some read from upward to downward. There is some concern that if the actual data plot is located in the top left of the lineup, observers may find it more easily than if it were placed in the bottom right. Ideally this does not happen. Experiment~9, conducted to test for structure in a real data set, enables this to be studied. Five different locations in the lineup were used to assess how fast and accurately observers identified the actual data plot.

@@

%\subsection{Sample of Null Plots} 
\item{\bf Sample of Null Plots:} In classical inference, the test statistic is compared with the full null distribution, to determine if it is extreme or not. In visual inference the actual plot is compared with plots of a finite number of samples from the null distribution. In the lineups used in the Turk experiments, there are 19 null plots upon which the comparison with the actual data plot is made. These null samples are random draws, and there is a chance that one or more null plots in the lineup may be similar or even more extreme than the actual data plot. The sample of null plots can affect the observer's decision. This is tested with the data from experiment~9, where the experiment was set up with lineups made from different null plots.

>>>>

In the initial Turk experiments (1,2,3) conducted with simulated data to compare the results from visual inference with those obtained by classical inference, the questions posed to the observers were more targeted. This was important \hh{to allow a direct comparison of the results to results from conventional testing. }%in order to make a comparison of the results. 
For latter studies (4, 7, 10), questions posed to the observer were more general. Table \ref{tbl:visual_stat} lists the questions used for each experiment.

@@

%\subsection{Learning Trend} 
\item{\bf Learning Trend:} One might expect that as an observer evaluates more lineups they may become more skillful in their evaluations. Each participant in  experiments~5, 6 and~7 was asked to evaluate a block of  ten lineups of the same type of data plot. The ten lineups were randomly chosen from the lineups  produced for each study. Before evaluating the first lineup, the observer needs to read instructions and become accustomed to the type of plot used in the lineup. In subsequent lineups  the type of plot is the same. It is possible that the observer becomes more skillful at recognizing the most different plot, either by more often detecting the actual data plot or more quickly reporting their choice. These two ways of measuring learning trend are evaluated on data collected from experiments~5, 6 and~7.

@@

%\subsection{Location of Actual Data Plot in the Lineup}
\item{\bf Location of Actual Data Plot in the Lineup} For all of the lineups used in the experiments a $5\times 4$ grid of 20 points is used. A random number generator is used to determine the position where the actual data plot is placed in each lineup. Eye tracking experiments \citep{zhao:2012} suggest that some observers traverse lineups in horizontal direction, while others have a vertical up-and-down approach to their search. Almost universal, observers start at the top left of a lineup. This leads to some concern that observers might be able to more easily identify the actual data plot, if it is placed at the top left of a lineup than when it the actual data is placed at the bottom right.
%With the help of eye-tracking equipment \citet{zhao:2012} tracked the observers' eyes to see their approach in searching the plots in a lineup. The results suggest that many people read lineups from left to right, some read from upward to downward. 
%There is some concern that if the actual data plot is located in the top left of the lineup, observers may be able to identify it more easily than when the actual data plot is at the bottom right. 
Ideally, this does not happen. Experiment~9  is designed to allow us to investigate the effects of location on detection. Five different locations in the lineup were used to assess how fast and accurately observers identified the actual data plot.

@@

%\subsection{Sample of Null Plots} 
\item{\bf Sample of Null Plots:} In classical inference, the test statistic is compared to the full null distribution, to decide if it is extreme or not. In visual inference the actual plot is compared to plots of a finite number of samples from the null distribution. In the lineups used in the Turk experiments, the actual data plot is compared on 19 null plots. These null samples are random draws, and there is a chance that one or more null plots in the lineup may be similar or even more extreme than the actual data plot. The sample of null plots can affect the observer's decision. This is tested with the data from experiment~9, where the experiment was set up with lineups made from different null plots.
